Supplementary Information -- Morphological convergence as on-line lexical analogy

### 2020-02-25 supplementary_information/si.Rnw
last round before Jen looks

test2 = test2 %>% 
  dplyr::mutate(
    verb_baseline_average = baseline_av - mean(baseline_av, na.rm = T), # center verb average
    participant_prereg_average = pre_reg_av - mean(pre_reg_av, na.rm = T), # center participant average
    bot_lexical_typicality = dplyr::case_when( # set factor levels used in paper, order factor
        lex_typicality == 'typical' ~ 'typical',
        lex_typicality == 'random' ~ 'random',
        lex_typicality == 'atypical' ~ 'reversed'
      ) %>% 
      factor(levels = c('typical', 'random', 'reversed')),
    bot_regularisation_rate = reg_rate %>% # order factor
      factor(levels = c('nc', '-40%', '+40%')),
    rescaled_CELEX_mgl_feature_distance = scales::rescale(baseline_mgl_features), # scale regular weights
    rescaled_CELEX_gcm_feature_distance = scales::rescale(baseline_gcm_features),
    rescaled_individual_mgl_feature_distance = scales::rescale(individual_mgl_features),
    rescaled_individual_gcm_feature_distance = scales::rescale(individual_gcm_features),
    gcm_score_diff = rescaled_individual_gcm_feature_distance - rescaled_CELEX_gcm_feature_distance, # calculate score diff
    mgl_score_diff = rescaled_individual_mgl_feature_distance - rescaled_CELEX_mgl_feature_distance
  )

@@

\noindent In this supplementary information, we detail (\ref{appendixverbs}) the structure of our nonce verb stimuli, the setup of (\ref{appendixgcm}) the Generalized Context Model (GCM) and (\ref{appendixmgl}) the Minimal Generalization Learner (MGL), and (\ref{appendixgcmvsmgl}) how these models compare. 
We illustrate the models using fits on our \emph {baseline} data.

@@

<<gcmtable, echo=F, eval=T, results=tex>>=
print(xtable::xtable(gcm_table, digits=2, label="gcmsetsize", caption="Number of forms in each verb class, GCM training data"))

@@

@@

<<mgltable, echo=F, eval=T, results=tex>>=
print(xtable::xtable(mgl_table, digits=2, label="mglsetsize", caption="Number of forms in each verb class, MGL training data"))

@@

As we see, however, the MGL also focuses the training set. The structural descriptions of the forms only cover a fraction of all verbs in the starting dictionary, organized by shared groups of segments. While disjunction increases the power of a theory (increasing the set of acceptable classes), disjunct classes frequently serve as input for phonological processes \citep{mielke2008emergence} . In the MGL, the disjunction emerges despite the lack of an initial specification: while one rule accounts for the specific behavior of a given form, multiple rules cover a given class. The number of regular and irregular rules posited by the MGL for each verb class can be seen in Table \ref{mglruleses}.

@@

<<mgltable, echo=F, eval=T, results=tex>>=
print(xtable::xtable(mgl_table, digits=2, label="mglruleses", caption="The number of unique rules posited by the MGL in each verb class"))

@@

lme.fit1 = glmer(resp_post_reg ~ 
    1 + verb_baseline_average * bot_lexical_typicality + participant_prereg_average + bot_regularisation_rate +
    (1 + verb_baseline_average | participant_id) + 
    (1  | word), 
  data = test2, 
  family = binomial (link = logit),
  control=glmerControl(optimizer="bobyqa")
)

@@

lme.fit1b = glmer(resp_post_reg ~ 
    1 + verb_baseline_average + bot_lexical_typicality + participant_prereg_average + bot_regularisation_rate +
    (1 + verb_baseline_average | participant_id) + 
    (1  | word), 
  data = test2, 
  family = binomial (link = logit),
  control=glmerControl(optimizer="bobyqa")
)

@@

lme.fit1c = glmer(resp_post_reg ~ 
    1 + verb_baseline_average * bot_lexical_typicality * bot_regularisation_rate + participant_prereg_average + 
    (1 + verb_baseline_average | participant_id) + 
    (1  | word), 
  data = test2, 
  family = binomial (link = logit),
  control=glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 100000))
)

@@

print(xtable::xtable(lme.fit1.sum, digits = 2, label = 'regressiontable1', caption = 'Experimental Data, Post-test Regression Model Summary. Model formula: post-test regular response ~ verb baseline mean x lexical typicality + participant pretest mean + regularization shift + (1 + verb.baseline.mean | participant) + (1 | verb)'))
@

@@

test2b = test2 %>% 
  dplyr::filter(
    bot_lexical_typicality != 'typical'
  )

fit_power1 = glmer(resp_post_reg ~ 
    1 + verb_baseline_average * bot_lexical_typicality + participant_prereg_average + bot_regularisation_rate +
    (1 + verb_baseline_average | participant_id) + 
    (1  | word), 
  data = test2b, 
  family = binomial ( link = logit ),
  control=glmerControl(optimizer="bobyqa")
  )

fit_power2 = glmer(resp_post_reg ~ 
    1 + verb_baseline_average + bot_lexical_typicality + participant_prereg_average + bot_regularisation_rate +
    (1 + verb_baseline_average | participant_id) + 
    (1  | word), 
  data = test2b, 
  family = binomial ( link = logit ),
  control=glmerControl(optimizer="bobyqa")
  )

@@

<<celex_resid1, echo=F,eval=T, results=tex>>=
load('models/si_celex_m3.rda')
celex_m3_summary = niceCoefTable(celex_m3)
print(xtable::xtable(celex_m3_summary, digits = 2, label = 'resid1', caption = 'Model 3: MGL scores and GCM residuals, baseline data'))
@

<<celex_resid2, echo=F,eval=T, results=tex>>=
load('models/si_celex_m4.rda')
celex_m4_summary = niceCoefTable(celex_m4)
print(xtable::xtable(celex_m4_summary, digits = 2, label = 'resid2', caption = 'Model 4: GCM scores and MGL residuals, baseline data'))
@

@@

<<indiv_diff1, echo=F, eval=T, results=tex>>=
load('models/si_indiv_m5.rda')
indiv_m5_sum = niceCoefTable(indiv_m5)
print(xtable::xtable(indiv_m5_sum, digits = 2, label = 'indiv1', caption = 'Model 5: Celex-GCM and individual score difference, post-test data'))

@@

<<indiv_diff2, echo=F,eval=T, results=tex>>=
load('models/si_indiv_m6.rda')
indiv_m6_sum = niceCoefTable(indiv_m6)
print(xtable::xtable(indiv_m6_sum, digits = 3, label = 'indiv2', caption = 'Model 6: Celex-MGL and individual score difference, post-test data'))
@

@@

<<comparison_model3, echo=F,eval=T, results=tex>>=
comp_fit1_sum = niceCoefTable(comp_fit1)
print(xtable::xtable(comp_fit1_sum, digits = 2, label = 'comp3', caption = 'Celex-MGL and Individual-GCM predict variation in the post-test'))
@

@@

# first, verb baseline regularisation averages and post-test averages across lex typicality
first_panel = test2 %>% 
  dplyr::group_by(
    word,
    bot_lexical_typicality,
    baseline_av # baseline average is there already
  ) %>%
  dplyr::summarise(
    post_reg_mean = resp_post_reg %>% # we get mean post test regularisation per lex typ
      mean()
  ) %>% 
  ungroup() %>% 
  dplyr::mutate( # we scale everything
    baseline = scales::rescale(baseline_av), # we call them baseline and posttest respectively
    posttest = scales::rescale(post_reg_mean),
    type = 'Participant responses' # we name this
  )

# same for gcm weights. 
second_panel = test2 %>% 
  dplyr::group_by(
    word,
    bot_lexical_typicality,
    rescaled_CELEX_gcm_feature_distance
  ) %>% 
  dplyr::rename(
    baseline = rescaled_CELEX_gcm_feature_distance
  ) %>% 
  dplyr::summarise(
    posttest = mean(rescaled_individual_gcm_feature_distance)
  ) %>% 
  dplyr::mutate(
    type = 'GCM'
  )

# same for MGL weights
third_panel = test2 %>% 
  dplyr::group_by(
    word,
    bot_lexical_typicality,
    rescaled_CELEX_mgl_feature_distance
  ) %>% 
  dplyr::rename(
    baseline = rescaled_CELEX_mgl_feature_distance
  ) %>% 
  dplyr::summarise(
    posttest = mean(rescaled_individual_mgl_feature_distance)
  ) %>% 
  dplyr::mutate(
    type = 'MGL'
  )

@@

# plot this, x is baseline, y is posttest.
# colour AND shape are lexical typicality.
# split panels across raw data / gcm / mgl.
# add loess smooths.
p1 = ggplot(three_studies, aes(x = baseline, y = posttest, colour = bot_lexical_typicality, shape = bot_lexical_typicality, linetype = bot_lexical_typicality)) +
  guides(size = F) +
  geom_point() +
  geom_smooth(se = F, method = 'loess', formula = 'y ~ x') +
  facet_wrap( ~ type) +
  scale_shape_discrete(name="Lexical\ntypicality") +
  # scale_colour_discrete(name="Lexical\ntypicality") +
  xlab('verb baseline regularization') +
  ylab('verb ESP regularization') +
  guides(linetype = F) +
  scale_colour_brewer(palette = 'Set2', name="Lexical\ntypicality")

>>>>

test2 = test2 %>% 
  dplyr::mutate(
    verb_baseline_average = baseline_av - mean(baseline_av, na.rm = T), # center verb average
    participant_prereg_average = pre_reg_av - mean(pre_reg_av, na.rm = T), # center participant average
    co_player_lexical_typicality = dplyr::case_when( # set factor levels used in paper, order factor
        lex_typicality == 'typical' ~ 'typical',
        lex_typicality == 'random' ~ 'random',
        lex_typicality == 'atypical' ~ 'reversed'
      ) %>% 
      factor(levels = c('typical', 'random', 'reversed')),
    co_player_regularisation_rate = reg_rate %>% # order factor
      factor(levels = c('nc', '-40%', '+40%')),
    rescaled_CELEX_mgl_feature_distance = scales::rescale(baseline_mgl_features), # scale regular weights
    rescaled_CELEX_gcm_feature_distance = scales::rescale(baseline_gcm_features),
    rescaled_individual_mgl_feature_distance = scales::rescale(individual_mgl_features),
    rescaled_individual_gcm_feature_distance = scales::rescale(individual_gcm_features),
    gcm_score_diff = rescaled_individual_gcm_feature_distance - rescaled_CELEX_gcm_feature_distance, # calculate score diff
    mgl_score_diff = rescaled_individual_mgl_feature_distance - rescaled_CELEX_mgl_feature_distance
  )

@@

In this supplementary information, we detail (\ref{appendixverbs}) the structure of our nonce verb stimuli, the setup of (\ref{appendixgcm}) the Generalized Context Model (GCM) and (\ref{appendixmgl}) the Minimal Generalization Learner (MGL), and (\ref{appendixgcmvsmgl}) how these models compare. We illustrate the models using fits on our \emph {baseline} data.

@@

<<gcmtable, echo=F, eval=T, results=tex>>=
print(xtable::xtable(gcm_table, digits=2, include.rownames = FALSE, rowlabel="gcmsetsize", caption="Number of forms in each verb class, GCM training data"), include.rownames = F)

@@

mgl_table = mgl_table %>% 
  mutate(
    category = toupper(category)
  )

@@

<<mgltable, echo=F, eval=T, results=tex>>=
print(xtable::xtable(mgl_table, digits=2, label="mglsetsize", caption="Number of forms in each verb class, MGL training data"), include.rownames = F)

@@

As we see, however, the MGL also focuses the training set. The structural descriptions of the forms only cover a fraction of all verbs in the starting dictionary, organized by shared groups of segments. While disjunction increases the power of a theory (increasing the set of acceptable classes), disjunct classes frequently serve as input for phonological processes \citep{mielke2008emergence} . In the MGL, the disjunction emerges despite the lack of an initial specification: while one rule accounts for the specific behavior of a given form, multiple rules cover a given class. The number of regular and irregular rules posited by the MGL for each verb class can be seen in Table \ref{mgltable}.

@@

@@

lme.fit1 = glmer(resp_post_reg ~ 
    1 + verb_baseline_average * co_player_lexical_typicality + participant_prereg_average + co_player_regularisation_rate +
    (1 + verb_baseline_average | participant_id) + 
    (1  | word), 
  data = test2, 
  family = binomial (link = logit),
  control=glmerControl(optimizer="bobyqa")
)

@@

lme.fit1b = glmer(resp_post_reg ~ 
    1 + verb_baseline_average + co_player_lexical_typicality + participant_prereg_average + co_player_regularisation_rate +
    (1 + verb_baseline_average | participant_id) + 
    (1  | word), 
  data = test2, 
  family = binomial (link = logit),
  control=glmerControl(optimizer="bobyqa")
)

@@

lme.fit1c = glmer(resp_post_reg ~ 
    1 + verb_baseline_average * co_player_lexical_typicality * co_player_regularisation_rate + participant_prereg_average + 
    (1 + verb_baseline_average | participant_id) + 
    (1  | word), 
  data = test2, 
  family = binomial (link = logit),
  control=glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 100000))
)

@@

print(xtable::xtable(lme.fit1.sum, digits = 2, label = 'regressiontable1', caption = 'Experimental Data, Post-test Regression Model Summary. Model formula: post-test regular response ~ verb baseline mean x lexical typicality + participant pretest mean + regularization shift + (1 + verb.baseline.mean | participant) + (1 | verb)'), include.rownames = F)
@

@@

test2b = test2 %>% 
  dplyr::filter(
    co_player_lexical_typicality != 'typical'
  )

fit_power1 = glmer(resp_post_reg ~ 
    1 + verb_baseline_average * co_player_lexical_typicality + participant_prereg_average + co_player_regularisation_rate +
    (1 + verb_baseline_average | participant_id) + 
    (1  | word), 
  data = test2b, 
  family = binomial ( link = logit ),
  control=glmerControl(optimizer="bobyqa")
  )

fit_power2 = glmer(resp_post_reg ~ 
    1 + verb_baseline_average + co_player_lexical_typicality + participant_prereg_average + co_player_regularisation_rate +
    (1 + verb_baseline_average | participant_id) + 
    (1  | word), 
  data = test2b, 
  family = binomial ( link = logit ),
  control=glmerControl(optimizer="bobyqa")
  )

@@

<<celex_resid1, echo=F,eval=T, results=tex>>=
load('models/si_celex_m3.rda')
celex_m3_summary = niceCoefTable(celex_m3)
print(xtable::xtable(celex_m3_summary, digits = 2, label = 'resid1', caption = 'Model 3: MGL scores and GCM residuals, baseline data'), include.rownames = F)
@

<<celex_resid2, echo=F,eval=T, results=tex>>=
load('models/si_celex_m4.rda')
celex_m4_summary = niceCoefTable(celex_m4)
print(xtable::xtable(celex_m4_summary, digits = 2, label = 'resid2', caption = 'Model 4: GCM scores and MGL residuals, baseline data'), include.rownames = F)
@

@@

<<indiv_diff1, echo=F, eval=T, results=tex>>=
load('models/si_indiv_m5.rda')
indiv_m5_sum = niceCoefTable(indiv_m5)
print(xtable::xtable(indiv_m5_sum, digits = 2, label = 'indiv1', caption = 'Model 5: Celex-GCM and individual score difference, post-test data'), include.rownames = F)

@@

<<indiv_diff2, echo=F,eval=T, results=tex>>=
load('models/si_indiv_m6.rda')
indiv_m6_sum = niceCoefTable(indiv_m6)
print(xtable::xtable(indiv_m6_sum, digits = 3, label = 'indiv2', caption = 'Model 6: Celex-MGL and individual score difference, post-test data'), include.rownames = F)
@

@@

<<comparison_model3, echo=F,eval=T, results=tex>>=
comp_fit1_sum = niceCoefTable(comp_fit1)
print(xtable::xtable(comp_fit1_sum, digits = 2, label = 'comp3', caption = 'Celex-MGL and Individual-GCM predict variation in the post-test'), include.rownames = F)
@

@@

# first, verb baseline regularisation averages and post-test averages across lex typicality
first_panel = test2 %>% 
  dplyr::group_by(
    word,
    co_player_lexical_typicality,
    baseline_av # baseline average is there already
  ) %>%
  dplyr::summarise(
    post_reg_mean = resp_post_reg %>% # we get mean post test regularisation per lex typ
      mean()
  ) %>% 
  ungroup() %>% 
  dplyr::mutate( # we scale everything
    baseline = scales::rescale(baseline_av), # we call them baseline and posttest respectively
    posttest = scales::rescale(post_reg_mean),
    type = 'Participant responses' # we name this
  )

# same for gcm weights. 
second_panel = test2 %>% 
  dplyr::group_by(
    word,
    co_player_lexical_typicality,
    rescaled_CELEX_gcm_feature_distance
  ) %>% 
  dplyr::rename(
    baseline = rescaled_CELEX_gcm_feature_distance
  ) %>% 
  dplyr::summarise(
    posttest = mean(rescaled_individual_gcm_feature_distance)
  ) %>% 
  dplyr::mutate(
    type = 'GCM'
  )

# same for MGL weights
third_panel = test2 %>% 
  dplyr::group_by(
    word,
    co_player_lexical_typicality,
    rescaled_CELEX_mgl_feature_distance
  ) %>% 
  dplyr::rename(
    baseline = rescaled_CELEX_mgl_feature_distance
  ) %>% 
  dplyr::summarise(
    posttest = mean(rescaled_individual_mgl_feature_distance)
  ) %>% 
  dplyr::mutate(
    type = 'MGL'
  )

@@

# plot this, x is baseline, y is posttest.
# colour AND shape are lexical typicality.
# split panels across raw data / gcm / mgl.
# add loess smooths.
p1 = ggplot(three_studies, aes(x = baseline, y = posttest, colour = co_player_lexical_typicality, shape = co_player_lexical_typicality, linetype = co_player_lexical_typicality)) +
  guides(size = F) +
  geom_point() +
  geom_smooth(se = F, method = 'loess', formula = 'y ~ x') +
  facet_wrap( ~ type) +
  scale_shape_discrete(name="Lexical\ntypicality") +
  # scale_colour_discrete(name="Lexical\ntypicality") +
  xlab('verb baseline regularization') +
  ylab('verb ESP regularization') +
  guides(linetype = F) +
  scale_colour_brewer(palette = 'Set2', name="Lexical\ntypicality")
